An investigation into the contentious phylogenetic results of the *Mixta* species

### 2021-04-17 Manuscript/Manuscript_Paper.Rmd
Update Manuscript_Paper.Rmd

---
title: "An investigation into the contentious phylogenetic results of the *Mixta* species"
author: "Kimberly N. Hinz, John Stavrinides*"
output:
  word_document:
    reference_docx: Style_RMarkdown.docx
    fig_caption: true
    # toc: true
    # toc_depth: 5
bibliography: References.bib
csl: IJSEM.csl
---
```{r Packages, echo = FALSE, message = FALSE, warning=FALSE}
library("knitr")
library("tidyverse")
library("ggplot2")
library("RColorBrewer")
library("extrafont")
theme_set(theme_classic())
```

@@

**Correspondence:*
John Stavrinides
Department of Biology
University of Regina
3737 Wascana Parkway
Regina, Saskatchewan
S4S0A2
Ph: (306) 337-8478
Email: john.stavrinides@uregina.ca

*Keywords*:

\newpage
# Abstract
*Mixta* is a recently described genus within the bacterial family *Erwiniaceae* and is closely related to *Pantoea*, *Erwinia*, and *Tatumella*. Depending on the genes and method used for phylogenetic analysis, the genus with which *Mixta* shares the most recent ancestor differs. This study aimed to determine the cause behind these contentious results and identified genes that may have resulted from horizontal gene transfer. Ten complete or partial genomes – two *Mixta* species, six species from other *Erwiniaceae* genera, and two outgroups – were retrieved from NCBI and annotated. Homologous genes were extracted yielding a dataset of 799 genes. The genes were aligned with the ClustalW algorithm, and `MEGA-CC` was used for model testing and computing phylogenetic trees for each gene for both nucleotide and amino acid sequences. Distance matrices were then extracted from these trees. Nucleotide and amino acid identity analyses were also done using the programming language `R`. *Pantoea* was the closest relative to the *Mixta* species in most analyses; however, results were not consistent. Some genes were also found to be more similar to other, non-*Pantoea* species. Diligence must be given to selecting genes for phylogenetic analysis and the method chosen to prevent any xenologous signal from distorting the actual relationships. Furthermore, future research should consider that different phylogenetic analyses may provide different results.

\newpage
# Introduction
The bacterial family Erwiniaceae includes diverse Gram-negative [@Winslow1917; @Prakash2015] organisms, including the genera *Mixta*, *Pantoea*, *Erwinia*, *Tatumella*, *Buchnera*, *Wigglesworthia*, *Phaseolibacter*, *Kalamiella* (LPSN, http://www.bacterio.net/). Numerous studies have focused on three of these genera – *Mixta*, *Pantoea*, and *Erwinia* – given their relevance to plant and human pathogens. For example, the genus *Erwinia* contains plant pathogens (e.g. *E. amylovora*) [@HaubenSwings2005; @Lopez2011; @Matsuura2012], epiphytes (e.g. *E. tasmaniensis*) [@Geider2006; @Mergaert1999], species that may contribute to plant disease [@Buonaurio2015], and others that are insect-associated [@Campillo2015; @Harada1997; @Skrodenyte-Arbaciauskiene2012]. Similarly, *Pantoea*, a genus reported by Gavini et al.[@Gavini1989], includes species that inhabit a variety of environments [@Gavini1989; @Popp2010; @Liu2013], including human and plant pathogens [@Prakash2015; @deBaere2004; @Fullerton2007; @Kratz2003; @Lim2006; @Schmid2003; @vanRostenberghe2006] as well as species associated with plant growth [@Smits2011; @Kim2012], insects [@Palmer2016], and fungi [@Ma2016; @Rong2016]. *Mixta*, originally *Pantoea*, is a novel genus first suggested by Palmer et al. [@Palmer2017] and later proposed by the same researchers in 2018 [@Palmer2018]. *M. calida* and *M. gaviniae* (previously species of *Pantoea*) were initially found in infant formula and its production environment [@Popp2010]. Fritz et al. [@Fritz2014] demonstrated that some strains of *M. calida* were linked with post-surgical meningitis in humans. Currently, there are only five *Mixta* species validly published (LPSN, http://www.bacterio.net/).

Understanding the phylogeny of bacterial species and genes allows researchers to investigate the species' evolutionary histories, relatedness, and the individual genes that enable them to survive under various conditions [@Palmer2017; @Bennett2012; @Prasanna2013; @Angus2014; @deMaayer2014; @Fouts2016]. The most common methods in phylogenetic analysis are those that use ubiquitous genes including 16S ribosomal RNA (rRNA) sequences [@Konstantinidis2007], multilocus sequence analysis (MLSA) with several housekeeping genes [@Konstantinidis2007; @Gevers2005; @Glaeser2015], and ribosomal MLSA with 53 structural ribosomal proteins [@Palmer2017; @Bennett2012; @Jolley2012]. However, the phylogenetic trees generated from these methods may not be robust [@Konstantinidis2007; @Glaeser2015; @Brady2008] because of the lower phylogenetic signal from the highly conserved genes [@Konstantinidis2007; @Gevers2005; @Fox1992; @Konstantinidis2005; @Staley2006; @RichterRosselloMora2009]. This issue is particularly the case with *Erwiniaceae* because much of our understanding is based on the 16S rRNA and other housekeeping genes [@Francino2006; @Hauben1998; @Naum2008; @Sproer1999; @Adeolu2016]. However, such analyses have an insufficient resolution at the species level [@Glaeser2015], and the genera *Pantoea*, *Erwinia*, *Tatumella*, and *Mixta* are often not monophyletic [@Rezzonico2016]. Another issue is the unintentional inclusion of paralogues (originating from an intragenomic duplication event) or xenologues (arising from horizontal gene transfer) in MLSA gene analyses [@Palmer2017; @Koonin2005].

The DNA gyrase subunit (gyrB), DNA-directed RNA polymerase subunit (rpoB), ATP synthase subunit (atpD), and initiation translation factor (infB) are more reliable than 16S rRNA gene sequences for determining intra- and intergeneric relationships [@Hedegaard1999; @Dauga2002; @Mollet1997; @Paradis2005] and a robust method for the Erwiniaceae family [@Brady2010]. The genera within Erwiniaceae are monophyletic when the analysis uses concatenated partial gyrB, rpoB, atpD, and infB gene sequences [@Brady2008; @Brady2009; @Moretti2011].

Nevertheless, despite efforts to find a suitable method for identifying the species tree for Erwiniaceae genera, *Mixta*’s position relative to *Pantoea*, *Erwinia*, and *Tatumella* vastly differs between studies [@Prakash2015; @Palmer2017; @Palmer2018; @Adeolu2016; @Brady2011; @Rezzonico2016; @Brady2012; @Gueule2015]. Not only do different gene combinations provide different results, but different algorithms and models for alignments and phylogenetic trees also contribute to the contentious results [@Naum2008; @Octavia2014]. For example, Prakash et al. [@Prakash2015], Palmer et al. [@Palmer2017], Rezzonico et al. [@Rezzonico2016], and Gueule et al. [@Gueule2015] all used gyrB, rpoB, atpD, and infB nucleotide gene sequences in their phylogenetic analyses. Prakash et al. conducted a neighbour-joining statistical method while the others used a maximum-likelihood (ML) method. Three of the studies – Rezzonico et al., Prakash et al., and Gueule et al. – placed the *Mixta* species within the *Pantoea* genera. In contrast. the analysis in Palmer et al. identified *Mixta* outside of the other three genera. Palmer et al. [@Palmer2018], Rezzonico et al. [@Rezzonico2016], and Brady et al. [@Brady2012] also ran analyses with the four MLSA genes as amino acid sequences. *Mixta* was found to be related closest to *Tatumella* by Rezzonico et al. and *Pantoea* by Palmer et al. Brady et al. suggested that *Mixta* was a part of the *Pantoea* genus and not separate. A separate study by Palmer et al. @Palmer2017 computed two ML trees; one used 1039 core amino acid sequences, and the other used 52 ribosomal MLSA nucleotide sequences. The 1039 core gene tree suggested *Pantoea* and *Tatumella* were sister taxa followed by two *Mixta* species. However, the ribosomal MLSA gene tree proposed *Pantoea* and *Tatumella* were sister taxa followed by *Erwinia* and then by *Mixta*.

Two possible causes may be behind the contentious results in phylogenetic analyses. The first is the methods; this includes the software and the applied model; two different phylogenetic tree models will provide different results even if both trees are ML and use the same genes and sequence type. The second potential cause is the materials: the gene sequences. The inclusion of different genes between analyses may give different results if one or more in the group is xenologous or has a different evolutionary rate. Xenologous genes need to be identified and removed from the dataset before phylogenetic analysis of gene sets since horizontal gene transfer (HGT) events frequently occur within bacteria and may occur between members of the same family. Sequence type may also affect results due to codon redundancy and codon usage bias. Additionally, there is no common phylogenetic tree model between nucleotide and amino acid sequences; therefore, evolutionary history is estimated differently between sequence types.

This paper aims to determine the cause behind the *Mixta* species' contentious phylogenetic results. Ten complete or partial genomes – two for each of *Mixta*, *Pantoea*, *Erwinia*, and *Tatumella* and two outgroups – were retrieved from a public database. Homologous genes were extracted from these genomes in both the nucleotide and amino acid sequences. Distance matrices were exported from the phylogenetic trees, and these were used to determine which species were most closely related to *Mixta* for each gene. Nucleotide and amino acid identity analyses were also conducted for each gene to compare the results between evolutionary history and sequence similarity. Phylogenetic trees were then done with either one gene or multiple genes to confirm results.

# Materials and Methods
## Genomes and Gene Sets
Ten publicly available full or partial genomes of the species' type strains were retrieved from the National Center for Biotechnology Information (NCBI). Each of the four *Erwiniaceae* genera had two representatives, and *Enterobacter cloacae* and *Pseudomonas syringae* are the outgroups (Table 1). *E. cloacae* belongs to the same order as *Erwiniacae* (*Enterobacterales*), and *P. syringae* belongs to the same class (*Gammaproteobacteria*).

The genomes were annotated with Prokka v1.14.1 [@Seemann2014] using default parameters, and homologous genes were extracted using GET_HOMOLOGUES software package [@Contreras-Moreira2013] with the bidirectional best-hit search algorithm using default parameters. GET_HOMOLOGUES extracted 954 homologous genes between the ten species.

```{r Table1, echo=FALSE, message=FALSE, warning=FALSE, fig.cap=TRUE}
strain <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Table1_Genomes.csv", stringsAsFactors = FALSE)

kable(strain, caption = "Table 1. Genomes used in this study.")
```

In `R` [@RCore2019], two filter functions were written to pass through the nucleotide sequence gene files. Since both nucleotide and amino acid sequences were analyzed, the filters were only applied to the nucleotide sequences. A list of the passed nucleotide gene files was used to isolate the corresponding amino acid gene files. The first filter kept genes files that contained ten, single-copy sequences that were approximately the same length. The shortest gene length allowed was 90% the length of the longest sequence in the gene file. For example, if the longest sequence in a gene file was 1000 bp, then the other sequences had to be at least 900 bp. Consequently, 741 gene files passed through this filter. Excluded from the dataset were 38 gene files because at least one species had a second copy and another 175 were excluded for not meeting the length requirements.

Since there may be an evolutionary reason why at least one of the sequences did not meet the length requires, a second filter was applied to the 175 gene files that did not meet the initial length requirements. This filter took each file at a time and separated the sequences into two groups. One group contained the sequences that were longer than the initial 90% length cutoff and the other contained the sequences between 80 and 90% the length of the longest gene in the file. If at least one of the sequences was below this new 80% cutoff, then the file was excluded. To pass through, both representatives of a genus had to be in the same group. An additional 58 genes were added to the working dataset, resulting in 799 genes. A list of the 799 gene names was used to grab the corresponding amino acid files.

## Gene Alignment
Gene alignments were in `R` using a ClustalW algorithm. The gene fasta files were read into `R` using the `readDNAStringSet()` function (package: Biostrings) [@Pages2020]. Alignments were done using the `msaClustalW()` function (package: msa) [@Bodenhofer2015] using default parameters and setting the maximum number of iterations to 100. The aligned sequences were then converted into a readable format using `msaConvert()` (package: msa)[@Bodenhofer2015] and written into a new fasta file using the `write.fasta()` function (package: seqinr) [@Charif2007]

## Distance Matrices
After alignments, model testing was done for each gene through `MEGA-CC` [@Kumar2012] for nucleotide and amino acid sequences. `MEGA-CC` is called in a command-line terminal along with a text file listing the file pathways to the 799 genes and a MAO file detailing the analysis preferences. Default parameters were used for model testing. The output was a comma-separated values (CSV) file for each gene. These CSV files were read into `R`, and the model for phylogenetic analysis was extracted based on the lowest Bayesian Information Criterion (BIC). Twelve and twenty-three unique models were used for nucleotide (Table 2) and amino acid sequences (Table 3) were needed to create phylogenetic gene trees.

@@

The gene file paths were grouped according to unique models. For each model, a text file was written containing the pathways. `MEGA-CC` [@Kumar2012] was called in a command-line terminal along with a file pathways text file and the corresponding MAO file detailing the unique model. This step was repeated for each model. The MAO files specified the use of a maximum likelihood statistical method, 500 bootstrap replications, the model, and rate patterns to create a phylogenetic tree for each gene. If a model required a Gamma distribution, the number of discrete Gamma categories was set to 5, which is the default.

Distance matrices were extracted from each of the phylogenetic trees and saved as Excel files. These files were then read into `R`. The genetic distances between *Mixta* and the other species were extracted for further analysis in `R`. Tables were created in `R Markdown` using the `kable()` function (package: knitr) [@Xie2020] and figures were created using the `ggplot()` function (package: ggplot2) [@Wickham2016].

## Nucleotide and Amino Acid Identity
Individual species sequences of genes were aligned to both of the corresponding *Mixta* sequences in turn. Alignments were done in `R` using the same methods as for gene alignment of the whole gene file. The aligned sequences were then split into the individual sites using the `str_split()` function (package: stringr) [@Wickham2019]. A function was created using base `R` code to calculate the number of identical sites between the two sequences over the total number of sites to give a similar identity percentage. Values that equal or are close to 100% are more identical to the *Mixta* sequences.

# Results and Discussion
After the two filters, 799 of 954 homologous genes remained. The genes included those encoding rRNA sequences, enzymes, transporters, and housekeeping genes and were scattered across the *M. calida* and *M. gaviniae* genomes.

## Genetic Distance
### Closest Relative from Genetic Distances
Genetic distance gives a numeric estimate of evolutionary change between species. The species that has the shortest genetic distance to a species of interest is considered the closest relative. For the following analysis, model testing was done first to estimate the best model for each gene. Next, individual phylogenetic gene trees were created, and it was from these that distance matrices were extracted. *P. septica* was the closest relative to 66.3% and 66.5% of the *M. calida* and *M. gaviniae* genes, respectively, for the nucleotide sequences (Figure 1). In comparison, only 42.7% and 42.8% of *M. calida* and *M. gaviniae* genes, respectively, are closely related to *P. septica* for the amino acid sequences (Figure 1), but this is still the majority. These results suggest that *P. septica* is the closest relative to *Mixta*, followed by *P. agglomerans*. In contrast, the *Mixta* genes are more closely related to the other species in the study than *P. septica* in the amino acid analysis (Figure 1). For instance, the percentage of genes related to *P. agglomerans* increased by 8.4 and 8.5% for *M. calida* and *M. gaviniae*, respectively (Figure 1). This difference may be due to a loss of phylogenetic signal when using amino acid sequences caused by codon degeneracy and having a third of available sites.

```{r Figure1, echo = FALSE, message = FALSE, warning=FALSE, fig.cap="Figure 1. The percentage of genes (out of 799) that were most closely related to any of the eight non-*Mixta* species using genetic distances. The species with the shortest genetic distance from both of the *Mixta* species is considered the closest relative. Blue bars are the nucleotide sequences of the genes with light blue bars representing the percentage of *M. calida* genes and dark blue bars representing *M. gaviniae*. Green bars are the corresponding amino acid sequences with light green representing *M. calida* genes and dark blue bars representing *M. gaviniae*. Approximately 66% of *M. calida* and *M. gaviniae* genes were most closely related to *P. septica* when using nucleotide sequences compared to only 43% when using amino acid sequences.", fig.width=6.5}
number_relatives <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure1_FirstRelativeMixtaNTAA.csv",
                             stringsAsFactors = FALSE) %>%
  mutate(Species = factor(Species, levels = c("P. septica", "P. agglomerans", "E. tasmaniensis", "E. amylovora", "T. ptyseos", "T. saanichensis",
                                              "E. cloacae", "P. syringae"), ordered = TRUE),
         Type = factor(Type, levels = c("calida_NT", "gaviniae_NT", "calida_AA", "gaviniae_AA"), ordered = TRUE))

@@

Unexpectedly, the number of *Mixta* genes that were more closely related to *E. cloacae* increased by 3.1% when amino acid sequences were used (Figure 1). The differences in relatedness for *Mixta* genes between nucleotide and amino acid sequences suggests that sequence type impacts the results, even when using the same genes.

If genetic distance infers the same closest relative for a *Mixta* gene for both nucleotide and amino acid analyses, these results may be more reliable. Thus, for genes where the closest relative is an *Erwinia*, *Tatumella*, or *Enterobacter* species both nucleotide and amino acid analyses, they may have resulted from a horizontal gene transfer (HGT) event rather than vertical evolution from a *Pantoea* species. No *Mixta* genes were closely related to *P. syringae* for either the nucleotide or amino acid sequences (Figure 1). This result confirms that *P. syringae* is a suitable outgroup for this analysis and suggests that any HGT events may be restricted to  *Enterobacterales*.

### Best Model vs Best Available Model
In `MEGAX` [@Kumar2018], there are fewer model options for genetic distance than for phylogenetic trees. However, model testing using this software tests all of the available models for phylogenetic tree analysis. Therefore, the recommended model based on the lowest BIC value may not be available for genetic distance analysis. GTR (General Time Reversible) and HKY (Hasegawa-Kishino-Yano) models, as well as the invariant sites (+I) parameter, are not available for genetic distance analysis. Genetic distances for the nucleotide sequences were estimated in `MEGA-CC` using the limited model options to determine if this would affect closest relative percentages. The model testing results were filtered to remove the models not available for genetic distances in `MEGAX`. Then distance matrices were estimated for each gene using the best available model based on the lowest BIC value. As a result, only four of twelve models were available with more than 87% of the genes now using either the Tamura 3-parameter with a Gamma parameter (T92_G) or the Tamura-Nei with a Gamma parameter (TN93_G) (Table 4). When not restricted by availability, only 58% of the genes required these models (Table 4). Furthermore, the best model was not available for approximately 30.3% of the genes (242 genes). The mean BIC difference between the best and best available models  was 2.65 (range: 0, 122.6). In this instance, 0 means that the best available model is the same as the best model. When these zeroes are ignored, the mean BIC difference was 13.8 (range: 0.1, 122.6). These differences may result in a less accurate estimation of genetic distance between species.

```{r Table4, echo = FALSE, message = FALSE, warning=FALSE, fig.cap=TRUE}
models <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Table4_GenDistModelsNT.csv", stringsAsFactors = FALSE)

colnames(models) <- c("Model", "Number of genes", "Number of genes, if available")

kable(models, caption = "Table 4. The number of genes requiring each model when using phylogenetic tree analysis vs genetic distance analysis in MEGAX for nucleotide sequences.")
```

The closest relatives to both *Mixta* species were extracted from the distance matrices. *P. septica* was estimated to be the closest relative to 12.6 and 14.4% more genes for *M. calida* and *M. gaviniae*, respectively, when the best available models were used (Figure 2). The next most considerable difference was for genes calculated to be most closely related to *P. agglomerans* (Figure 2). Approximately 5.5 and 7.1% fewer genes were most closely related to *P. agglomerans* when the best available model was used (Figure 2). Additionally, there was more disparity in the closest relatives between *M. calida* and *M. gaviniae* when only the best available model was used (Figure 2). When the best model was used, the two *Mixta* typically had the same closest relative for each gene; however, this was not the case when model options were limited to those available. Due to differences in evolutionary change estimation between models and the addition or exclusion of additional parameters, genetic distances may be inaccurately calculated when the best model is not used. Such results may even suggest that sibling taxa are not as closely related as initially hypothesized.

@@

About 20.0 and 22.3% of *M. calida* and *M. gaviniae* genes, respectively, had a different closest relative between the best and the best available models. Thus, of the 30.3% of genes with different models between the two analyses, 10.3% of the *M. calida* genes and 8.0% of the *M. gaviniae* genes had the same closest relative. These genes' results may be the same because the two models were similar and explained approximately the same amount of variance. Genetic distances between the species may also be sufficiently large that it does not change which species is most closely related to *Mixta*.

Each model estimates genetic distance differently. The best model explains most of the variance between sequences and gives a more accurate estimation of genetic distance. Using a model that does not explain most of the variance will estimate genetic distance less accurately, resulting in related taxa (such as those from the same species) having a longer genetic distance between them than in actuality. This may also result in the opposite for taxa that are not closely related (i.e. *Pantoea* spp. and *E. cloacae*). Even though only two nucleotide models and one parameter were unavailable for genetic distance analysis in `MEGAX`, the results for the closest relative to *Mixta* genes were substantially impacted. This was particularly the case between the two *Mixta* as one would expect them to have the same closest relative for most of the homologous genes. Therefore, using the best model is recommended. However, this may not be possible when estimating genetic distance using more than one gene that may require different models. Thus, it must be considered that one or more the genes in the concatenation may have a different evolutionary pattern than the others and results should be investigated further to ensure accuracy.

Although genetic distance estimation in `MEGAX` did not offer all of the models and parameters suggested by model testing, all of the models and parameters were available for phylogenetic tree analysis and genetic distance can be exported from these trees. However, this has to be done manually by opening the tree file in `MEGAX` and exporting the distance matrix as either an Excel or CSV file. Therefore, it is not feasible when hundreds or thousands of genetic distance matrices are needed.

### Spatial Patterns of Closest Relatives
Circular plots were generated using the *Mixta* gene IDs for nucleotide (Figure 3) and amino acid sequences (Figure 4). These were created to assess whether there were any positional effects of the genes and spatial patterns in the closest relative. The gene IDs give the approximate location of each gene relative to the *Mixta* genome. The length and colour of a bar represent the species that was the closest relative to the *Mixta* gene at that position. Since the closest relatives were the same both *Mixta* species except for 1 and 2 genes for nucleotide and amino acid sequences, respectively, only the results for *M. calida* are shown. These genes will be discussed later on.

@@

While no distinct patterns emerged, genes most similar to *P. septica* predominate the *M. calida* genome (Figures 6 and 7). This suggests that *Mixta* and *P. septica* may be related as their sequences are similar and few mutations have occurred after they diverged.

>>>>

---
title: "An investigation into the contentious phylogenetic results of the *Mixta* species"
output:
  word_document:
    reference_docx: Style_Manuscript.docx
    fig_caption: true
    # toc: true
    # toc_depth: 5
bibliography: References.bib
csl: IJSEM.csl
---
```{r Packages, echo = FALSE, message = FALSE, warning=FALSE}
library("knitr")
library("tidyverse")
library("ggplot2")
library("RColorBrewer")
library("extrafont")
theme_set(theme_classic())
```

Kimberly N. Hinz, John Stavrinides^1^

@@

^1^**Correspondence**:  
John Stavrinides  
Department of Biology  
University of Regina  
3737 Wascana Parkway  
Regina, Saskatchewan  
S4S0A2  
Ph: (306) 337-8478  
Email: john.stavrinides@uregina.ca

**Keywords:**

<!-- \newpage -->
# Abstract
<!-- *Mixta* is a recently described genus within the bacterial family *Erwiniaceae* and is closely related to *Pantoea*, *Erwinia*, and *Tatumella*. Depending on the genes and method used for phylogenetic analysis, the genus with which *Mixta* shares the most recent ancestor differs. This study aimed to determine the cause behind these contentious results and identified genes that may have resulted from horizontal gene transfer. Ten complete or partial genomes – two *Mixta* species, six species from other *Erwiniaceae* genera, and two outgroups – were retrieved from NCBI and annotated. Homologous genes were extracted yielding a dataset of 799 genes. The genes were aligned with the ClustalW algorithm, and MEGA-CC was used for model testing and computing phylogenetic trees for each gene for both nucleotide and amino acid sequences. Distance matrices were then extracted from these trees. Nucleotide and amino acid identity analyses were also done using the programming language `R`. *Pantoea* was the closest relative to the *Mixta* species in most analyses; however, results were not consistent. Some genes were also found to be more similar to other, non-*Pantoea* species. Diligence must be given to selecting genes for phylogenetic analysis and the method chosen to prevent any xenologous signal from distorting the actual relationships. Furthermore, future research should consider that different phylogenetic analyses may provide different results. -->

# Intro and MM
<!-- \newpage -->
<!-- # Introduction -->
<!-- The bacterial family *Erwiniaceae* includes diverse Gram-negative [@Winslow1917; @Prakash2015] organisms, including the genera *Mixta*, *Pantoea*, *Erwinia*, *Tatumella*, *Buchnera*, *Wigglesworthia*, *Phaseolibacter*, and *Kalamiella* [REF CRAIG]. *Mixta* is a relatively newer genus first identified by Palmer et al. in 2017 [@Palmer2017] and later described by the same researchers in 2018 [@Palmer2018]. Currently, there are only five *Mixta* species validly published [@LPSN]. *M. calida* and *M. gaviniae* (previously species of *Pantoea*) were initially found in infant formula and its production environment [@Popp2010]. Fritz et al. [@Fritz2014] demonstrated that some strains of *M. calida* were linked with post-surgical meningitis in humans. The other *Mixta* species include *M. theicola* which was isolated from black tea extract [@Kato2015], *M. intestinalis* which was isolated from a healthy human fecal sample [@Prakash2015], and *M. tenebrionis* which was isolated from the gut microbiota of a plastic-eating mealworm [@Xia2020]. The first two were previously named as *Pantoea* species. -->

<!-- *Mixta*'s position relative to other members of the *Erwiniaceae*, in particular to *Pantoea*, *Erwinia*, *Tatumella*, differs vastly between studies [@Prakash2015; @Palmer2017; @Palmer2018; @Adeolu2016; @Brady2011; @Rezzonico2016; @Brady2012; @Gueule2015]. Prakash et al. [@Prakash2015], Palmer et al. [@Palmer2017], Rezzonico et al. [@Rezzonico2016], and Gueule et al. [@Gueule2015] all used *gyrB*, *rpoB*, *atpD*, and *infB* nucleotide gene sequences in their phylogenetic analyses. Prakash et al. [@Prakash2015] conducted a neighbour-joining statistical method while the others used a maximum-likelihood (ML) method [@Palmer2017; @Rezzonico2016; @Gueule2015]. Three of the studies – Rezzonico et al. [@Rezzonico2016], Prakash et al. [@Prakash2015], and Gueule et al. [@Gueule2015] – placed the *Mixta* species within the *Pantoea* genera. In contrast, the analysis in Palmer et al. [@Palmer2017] identified *Mixta* outside of the other three genera. A separate study by Palmer et al. [@Palmer2018], Rezzonico et al. [@Rezzonico2016], and Brady et al. [@Brady2012] also ran analyses with the four MLSA genes as amino acid sequences. *Mixta* was found to be related closest to *Tatumella* by Rezzonico et al. [@Rezzonico2016] and *Pantoea* by Palmer et al. [@Palmer2018]. Brady et al. [@Brady2012] suggested that *Mixta* was a part of the *Pantoea* genus and not separate. Additionally, Palmer et al. [@Palmer2017] computed two ML trees; one used 1039 core amino acid sequences, and the other used 52 ribosomal MLSA nucleotide sequences. The 1039 core gene tree suggested *Pantoea* and *Tatumella* were sister taxa followed by two *Mixta* species. However, the ribosomal MLSA gene tree proposed *Pantoea* and *Tatumella* were sister taxa followed by *Erwinia* and then by *Mixta*. -->

<!-- What remains unclear is the relative contribution of disaparate evolutionary histories of *Mixta* genes, tree-building methods, and sequence evolution on the phylogenetic position of the genus. This paper aims to determine the cause behind the *Mixta* species' contentious phylogenetic results. Ten complete or partial genomes – two for each of *Mixta*, *Pantoea*, *Erwinia*, and *Tatumella* and two outgroups – were retrieved from a public database. Homologous genes were extracted from these genomes in both the nucleotide and amino acid sequences. Distance matrices were exported from the phylogenetic trees, and these were used to determine which species were most closely related to *Mixta* for each gene. Nucleotide and amino acid identity analyses were also conducted for each gene to compare the results between evolutionary history and sequence similarity. Phylogenetic trees were constructed for combinations of different gene sequences.  -->

<!-- # Materials and Methods -->
<!-- ## Genomes and Gene Sets -->
<!-- Ten publicly available full or partial genomes of the species' type strains were retrieved from the National Center for Biotechnology Information (NCBI). Two representative genomes were used for each of the four *Erwiniaceae* genera while *Enterobacter cloacae* (*Enterobacterales*) and *Pseudomonas syringae* (*Pseudomonaceae*) were used as outgroups (Table 1). The genomes were annotated with Prokka v1.14.1 [@Seemann2014] using default parameters, and homologous genes were extracted using GET_HOMOLOGUES software package [@Contreras-Moreira2013] with the bidirectional best-hit search algorithm using default parameters. GET_HOMOLOGUES extracted 954 homologous genes between the ten species. -->

<!-- ```{r Table1, echo=FALSE, message=FALSE, warning=FALSE, fig.cap=TRUE} -->
<!-- strain <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Table1_Genomes.csv", stringsAsFactors = FALSE) -->

<!-- colnames(strain)[3] <- "GenBank Accession No." -->

<!-- kable(strain, caption = "Table 1. Genomes used in this study.") -->
<!-- ``` -->

<!-- In `R` [@RCore2019], two filter functions were written for the nucleotide sequence gene files. The first filter kept genes files that contained ten, single-copy sequences that were approximately the same length. The shortest gene length allowed was 90% the length of the longest sequence in the gene file. Of the 954 homologous gene sets, 38 multi-copy genes were excluded, and another 175 were excluded for not meeting the length requirements.  -->

<!-- A second filter was applied to these 175 genes to identify any genes wherein both representatives of each genera were approximately the same length. The filter split the genes into two groups: sequences that met the original 90% length cutoff and sequences that were between 80 and 90% the length of the longest sequence. If both representatives of a genus were in the same group, then the gene was added to the working dataset. If at least one of the sequences did not meet the 80% cutoff, then the gene was excluded. An additional 58 genes were added to the working dataset, resulting in 799 genes. The amino acid sequences for these genes were subsequently downloaded. -->

<!-- ## Gene Alignment -->
<!-- Gene alignments were done in `R` using a ClustalW algorithm. The fasta files were read into the session using the `readDNAStringSet()` function (package: Biostrings) [@Pages2020]. The function called to do the alignments was the `msaClustalW()` function (package: msa) [@Bodenhofer2015] using default parameters and 100 iterations. The aligned sequences were then converted into a comma-separated values (CSV)-compatible format using `msaConvert()` (package: msa)[@Bodenhofer2015] and written into a new fasta file using the `write.fasta()` function (package: seqinr) [@Charif2007] -->

<!-- ## Model Selection -->
<!-- The best model of sequence evolution was identified for each nucleotide and amino acid alignment using Model Selection as implemented in MEGA-CC [@Kumar2012]. Default parameters were used for model testing. The output was a CSV file for each gene. These CSV files were read into `R`, and the model for phylogenetic analysis was extracted based on the lowest Bayesian Information Criterion (BIC). -->

<!-- ## Phylogenetic Analysis -->
<!-- Phylogenetic analysis was performed in MEGA-CC using maximum likelihood with 500 bootstrap replications, the model, and rate patterns to create a phylogenetic tree for each gene. If a model required a Gamma distribution, the number of discrete Gamma categories was set to 5, which is the default. -->

<!-- Distance matrices were extracted from each of the phylogenetic trees and saved as Excel files. These files were then read into the `R` session. The genetic distances between *Mixta* and the other species were extracted for further analysis. Tables were created in `R Markdown` using the `kable()` function (package: knitr) [@Xie2020] and figures were created using the `ggplot()` function (package: ggplot2) [@Wickham2016]. -->

<!-- ## Nucleotide and Amino Acid Identity -->
<!-- Each of the non-*Mixta* sequences were aligned to one of the corresponding *Mixta* sequences in turn. Alignments were performed in R as described above, and the aligned sequences were then split into the individual sites using the `str_split()` function (package: stringr) [@Wickham2019]. A function was created using base `R` code to calculate the number of identical sites between each sequence and the *Mixta* homologs. Values that equal or are close to 100% are more identical to the *Mixta* sequences. -->

# Results
Twelve and twenty-three unique models were used for nucleotide (Table 2) and amino acid sequences (Table 3) were needed to create phylogenetic gene trees.

@@

After the two filters, 799 of 954 homologous genes remained. The genes included those encoding rRNA sequences, enzymes, transporters, and housekeeping genes and were scattered across the *M. calida* and *M. gaviniae* genomes.

## Best Model vs Best Available Model
In MEGAX [@Kumar2018], the user can either calculate genetic distance or estimate a phylogenetic tree. However, although model testing accesses all possible models available in the software, all models are not available for genetic distance analysis. The GTR (General Time Reversible) and HKY (Hasegawa-Kishino-Yano) models and the +I (invariant sites) parameter are available only for phylogenetic tree analysis.

```{r Figure1, echo = FALSE, message = FALSE, warning=FALSE, fig.cap="Figure 1. The percent of genes (n = 799) requiring each model when using the recommended model for phylogenetic tree analysis (green) and when using the best available model for genetic distance analysis (orange). Models are for nucleotide sequences.", fig.width=6.5}
models <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure1_BestvAvailableModels.csv", stringsAsFactors = FALSE)

ggplot(data = models, aes(x = Model, y = Percent, fill = Bst_Avlb)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_fill_brewer(palette = "Dark2",
                    labels = c("Phylogenetic", "Genetic Distance")) +
  labs(y = "Percent of Genes \n Requiring Each Model",
       fill = "Analysis") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),                                # Rotates x axis labels
        legend.position = c(0.3, 0.95),                                                   # Moves legend inside the plot and in the top-left corner
        legend.justification = c("right", "top"),                                         # Top right corner is at above coordinates
        legend.text.align = 0,                                                            # Aligns legend text to the left
        legend.margin = margin(6, 6, 6, 6),                                               # Margins around legend
        text = element_text(size = 12,  family = "Times New Roman"))
```

```{r Figure2, echo = FALSE, message = FALSE, warning=FALSE, fig.cap="Figure 2. The percentage of genes (n = 799) most closely related to any of the eight non-*Mixta* species. Closest relative was inferred using shortest genetic distance. Blue bars represent the nucleotide sequences and green bars represent the amino acid sequences. The lighter of each colour represents *M. calida* and the darker *M. gaviniae*.", fig.width=6.5}
number_relatives <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure2_FirstRelativeNTAA.csv",
                             stringsAsFactors = FALSE) %>%
  mutate(Species = factor(Species, levels = c("P. septica", "P. agglomerans", "E. tasmaniensis", "E. amylovora", "T. ptyseos", "T. saanichensis",
                                              "E. cloacae", "P. syringae"), ordered = TRUE),
         Type = factor(Type, levels = c("calida_NT", "gaviniae_NT", "calida_AA", "gaviniae_AA"), ordered = TRUE))

@@

@@

@@
